---
title: "Inequalities in PISA scores"
author: "D. Moreno and J. Moreno"
date: ""
output: html_document
---

```{r, ,echo=FALSE}
library(ggplot2)
library(RColorBrewer)
library(ineq)
library(plyr)
library(reshape2)
library(gridExtra)
```

This document was inspired by [this post](http://freakonometrics.hypotheses.org/16449) by [Arthur Charpentier](https://twitter.com/freakonometrics). 

The repository with the code for creating this post is [here](https://github.com/finiterank/pisa). Just some plots so far. We may try something more elaborated (e.g., explicit comparisons of Colombia with other countries) later on.

The data file is available [here](http://beta.icm.edu.pl/PISAcontest/data/student2012.rda). 

```{r, ,echo=FALSE}
load("data/student2012.rda")
st <- student2012
rm(student2012)
latin.america <- c("Colombia", "Brazil", "Chile", "Argentina", "Peru", "Costa Rica", "Mexico", "Uruguay", "United States of America")
```

# Entropy

One possible measure of inequality is entropy, the classical notion developed by Shannon for Information Theory. [This](http://en.wikipedia.org/wiki/Generalized_entropy_index) is a generalized version. I am using here the one with $\alpha=1$. 

```{r, echo=FALSE,  fig.height=9, fig.width=8, fig.align='center'}

#function.entropy <- function(x){
#  mx <- mean(x)
#  a <- x * (1/mx)
#  b <- log(a) * a
#  output <- mean(b)
#  return(output)
#}

tabla.entropy <- ddply(st, .(CNT, ST04Q01), summarise, entropy.math=entropy(PV1MATH, parameter=1), entropy.read=entropy(PV1READ, parameter=1), mean.math=mean(PV1MATH), mean.read=mean(PV1READ))
tabla.entropy$CNT <- reorder(tabla.entropy$CNT, tabla.entropy$entropy.math)
color.label.entropy <- ifelse(levels(tabla.entropy$CNT) %in% latin.america, "red", "black")
tabla.entropy.melt <- melt(tabla.entropy, id.vars=c("CNT", "ST04Q01"), measure.vars = c("entropy.math", "entropy.read"))
levels(tabla.entropy.melt$variable) <- c("Math", "Reading")
ggplot(tabla.entropy.melt, aes(x=CNT, y=value, fill=ST04Q01)) + geom_bar(stat="identity", position="dodge") + facet_grid(. ~ variable) + coord_flip() + xlab("Country") + ylab("Entropy (Theil index)") + theme(axis.text.y = element_text(colour=color.label.entropy, hjust=1)) + scale_fill_discrete(name="Sex")
```

Although entropy is not necessarily a measure of variance, collections of test scores with higher entropy tend to have lower average than those with lower entropy:

```{r, echo=FALSE, ,  fig.height=7, fig.width=7, fig.align='center'}
ggplot(tabla.entropy, aes(x=mean.math, y=entropy.math, col=ST04Q01)) + geom_point() + geom_smooth(method="lm") + xlab("Mean Math Scores") + ylab("Entropy Math Scores") + scale_color_discrete(name="Sex")
```

By the way, why is the entropy of female scores (almost) consistently lower than that of males of the same country? 

# Standard Deviation

Standard deviation seems to be the most commonly used indicator of performance inequality in standarized test scores. The ranking changes drastically: 

```{r, echo=FALSE, ,  fig.height=9, fig.width=8, fig.align='center'}
tabla.sd <- ddply(st, .(CNT, ST04Q01), summarise, sd.math=sd(PV1MATH), sd.read=sd(PV1READ))
tabla.sd$CNT <- reorder(tabla.sd$CNT, tabla.sd$sd.math)
color.label.sd <- ifelse(levels(tabla.sd$CNT) %in% latin.america, "red", "black")
tabla.sd <- melt(tabla.sd, id.vars=c("CNT", "ST04Q01"), measure.vars = c("sd.math", "sd.read"))
ggplot(tabla.sd, aes(x=CNT, y=value, fill=ST04Q01)) + geom_bar(stat="identity", position="dodge") + facet_grid(. ~ variable) + coord_flip()  + scale_fill_discrete(name="Sex") + theme(axis.text.y = element_text(colour=color.label.sd, hjust=1))
```

# Distributions of scores (for selected countries)

```{r, echo=FALSE}
paises <- c("Spain", "Canada",  "Israel", "Japan", "Germany", "Finland", "Estonia", "Ireland", "Qatar", "France", latin.america)
math <- st[(st$CNT %in% paises),c("CNT","ST04Q01", "PV1MATH", "PV1READ")]
```

First, a violin plot of distributions of scores in math (differentiating by sex and ordered by entropy):

```{r, echo=FALSE, ,  fig.height=14, fig.width=8, fig.align='center'}
math$CNT <- factor(math$CNT, levels=levels(tabla.entropy$CNT)[levels(tabla.entropy$CNT) %in% paises])
color.label <- ifelse(levels(math$CNT) == "Colombia", "red", "black")
ggplot(math, aes(x=CNT, y=PV1MATH)) + 
  geom_violin(aes(fill=factor(CNT == "Colombia"))) + 
  facet_grid(. ~ ST04Q01) + 
  xlab("Country") + 
  ylab("Math Score") +
  scale_fill_brewer(guide = 'none', palette="Set1") + 
  coord_flip() + theme(axis.text.y = element_text(colour=color.label, hjust=1))
```

Empirical cummulative density function for each country in math:

```{r, echo=FALSE, ,  fig.height=14, fig.width=7, fig.align='center'}
#colombia.male.ecdf<- ecdf(math[(math$CNT=="Colombia") & (math$ST04Q01 == "Male"), "PV1MATH"])

ggplot(math, aes(x=PV1MATH, colour=ST04Q01)) + stat_ecdf() + facet_grid(CNT ~ .) + scale_colour_brewer(guide = 'none', palette="Set1") + xlab("Math Score") + ylab("Cummulative Density")
```

And a kernel density estimate (also for math):

```{r, echo=FALSE, ,  fig.height=14, fig.width=7, fig.align='center'}
ggplot(math, aes(x=PV1MATH, colour=ST04Q01)) + geom_density() + facet_grid(CNT ~ .) + scale_colour_brewer(guide = 'none', palette="Set1") + xlab("Math Score") + ylab("Density")
```

# Quantiles

In his post, Charpentier compares France's score distribution with other countries by plotting the difference of the quantiles at each level. Here I do the same for Colombia against 17 other countries with the math scores. Just as an illustration I also include Colombia versus Colombia (red is male and blue is female).

```{r, echo=FALSE, fig.height=14, fig.width=9, fig.align='center'}
pisa.math.scores <- function(cnt, sex){
  return(st[(st$CNT == cnt) & (st$ST04Q01 ==sex),"PV1MATH"])
}

dif.quant.scores <- function(cntA, cntB, sex, interval){
  cntA.quant <- quantile(pisa.math.scores(cntA, sex), probs = interval)
  cntB.quant <- quantile(pisa.math.scores(cntB, sex), probs = interval)
  return(cntA.quant - cntB.quant)
  }

plot.quantiles <- function(cntA, cntB){
  interval <- seq(0,1, 0.001)
  dif.quant.m <- dif.quant.scores(cntA, cntB, "Male", interval)
  dif.quant.f <- dif.quant.scores(cntA, cntB, "Female", interval)
  df <- data.frame(interval, dif.quant.m, dif.quant.f)
  df <- melt(df, id.vars = "interval")
  ggplot(df, aes(x=interval, y=value, col=variable)) + 
    geom_smooth(method="loess") +
    scale_color_discrete(guide="none", name="Sex", labels=c("Male", "Female")) +
    xlab("Probability") + 
    ylab("Quantile") + ggtitle(paste(cntA, " - ", cntB, sep="")) +
    theme(plot.title = element_text(size = rel(.8)), 
          plot.margin = unit(c(0,.2,.2,.2), "lines"),
          axis.title.x = element_blank(),
          axis.title.y = element_blank())
  }

paises.comp <- c(latin.america, "Japan", "Finland", "Canada", "Spain", "Germany", "Estonia", "France", "Portugal", "Greece")

plots.quant <- lapply(paises.comp, function(x) plot.quantiles("Colombia", x))

do.call(grid.arrange, c(plots.quant, ncol=3))
```

And let's do the same with Singapore:

```{r, echo=FALSE, fig.height=14, fig.width=9, fig.align='center'}
plots.quant.singapore <- lapply(paises.comp, function(x) plot.quantiles("Singapore", x))

do.call(grid.arrange, c(plots.quant.singapore, ncol=3))
```


# Further reading

Serious approaches to this problem:

* [The Measurement of Educational Inequaility](http://elibrary.worldbank.org/doi/pdf/10.1596/1813-9450-5873), by Francisco H. G. Ferreira and Jérémie Gignoux
* [Equality of educational opportunity employing PISA data](http://www.urosario.edu.co/urosario_files/f9/f9b47c44-b3d4-49e6-a4c7-cb9b82a2953c.pdf), by Márcia de Carvalho, Luis Fernando Gamboa and Fábio D. Waltenberg